Equalizer\\Programming and User Guide

### 2010-12-21 Developer/ProgrammingGuide/paper.tex
Update Quickstart Guide Update Programming Guide: - Collage-related changes - Inline code from Eq directly, instead of cpy-pasting it - General review for 1.0a - related code cleanups
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications using
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.7.9 for Equalizer 0.9.1+svn\\[\medskipamount]
%  \today
}

@@

\lstset{language=C++}
\section*{Equalizer 0.9.1 Programming and User Guide}
\today
\subsection*{Contributors}

@@

The images on the front page show the following applications build using
Equalizer: 
RTT DeltaGen\footnote{Image copyright Realtime Technology AG, 2008} [top
right], 
\textsf{eqRaster}, a scalable terrain visualization\footnote{Image
  courtesy of Visualization and Multimedia Lab, University of Zurich}
[middle center],
a flow visualization application for climate research\footnote{Image
  courtesy of Computer Graphics and Multimedia Systems, University of
  Siegen} [middle right],
the \textsf{eqPly} polygonal renderer in a three-sided CAVE [bottom
left],
the \textsf{eVolve} volume renderer\footnote{Data set courtesy of
  General Electric, USA} [bottom center],
and the \textsf{eqNBody} gravitational n-body simulation, based on the
N-Body CUDA example [bottom right].

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
    1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
    1.7.2   & Nov 18, 2009 & Add \sref{sSubpixel} and \sref{sFSAA} for Subpixel compounds\\
    1.7.3   & Feb 02, 2010 & Revision for Equalizer 0.9.1\\
    1.7.4   & Feb 22, 2010 & Add \sref{sSlaveCommit}\\
    1.7.5   & Jun 03, 2010 & Add \sref{sClusterSetup}\\
    1.7.6   & Aug 23, 2010 & General update for fabric changes\\
    1.7.7   & Oct 01, 2010 & SystemPipe and SystemWindow renaming\\
    1.7.8   & Oct 04, 2010 & Document GPU affinity workaround hint\\
    1.7.9   & Nov 03, 2010 & Add \sref{sErrorHandling}\\
    \hline
  \end{tabularx}}
\clearpage

@@

Equalizer is the standard middleware for the development and deployment
of parallel OpenGL applications. It enables applications to benefit from
multiple graphics cards, processors and computers to scale rendering
performance, visual quality and display size. An Equalizer-based
application runs unmodified on any visualization system, from a simple
workstation to large scale graphics clusters, multi-GPU workstations and
Virtual Reality installations.

@@

Equalizer provides the domain-specific parallel rendering know-how and
abstracts configuration, threading, synchronization, windowing and event
handling. It is a `GLUT on steroids', providing parallel and distributed
execution, scalable rendering features, network data distribution and
fully customizable event handling.

@@

Compiling Equalizer is as simple as running \textsf{make} on Linux or
Mac OS X, building the provided XCode project on Mac OS X 10.5, or
building the Equalizer Visual Studio 2005 solution on Windows. Note that
on Mac OS X 10.4 (Tiger), some prerequisites have to be installed before
running \textsf{make}, as explained in \textsf{README.Darwin}.

@@

The Equalizer architecture is based on a client-server model. The client
library exposes all functionality discussed in this document to the
programmer, and provides communication between the different Equalizer
processes. The underlying network layer uses a peer-to-peer structure,
allowing render clients to communicate directly with each other, e.g.,
during parallel compositing.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/processes.pdf}
  {\caption{\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The application can also be a rendering client, in which case it can
also contribute to the rendering. If it does not implement any render
client code, it is reduced to be the application's `master' process
without any OpenGL windows and 3D rendering.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
  {\caption{\label{fStereoCmp}\small A Stereo Compound}}
\end{wrapfigure}
Stereo compounds, as shown in \fig{fStereoCmp}\footnote{3D model
  courtesy of Stereolithography Archive at Clemson University.}, assign
each eye pass to individual rendering units. The resulting images are
copied to the appropriate stereo buffer. This mode supports a variety of
stereo modes, including active (quad-buffered) stereo, anaglyphic stereo
and auto-stereo displays with multiple eye passes.

@@

Stereo compounds are used by all applications when rendering in stereo,
and is often combined with other modes.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Subpixel.pdf}
  {\caption{\label{fSubpixel}\small A Subpixel Compound}}
\end{wrapfigure}
Applications can use subpixel compounds to accelerate depth-of-field effects and
software anti-aliasing, with potentially a different number of idle or non-idle
samples per pixel.

@@

Subpixel compounds are configured using the \textsf{subpixel} parameter, using
the values \textsf{[ index size ]}. The \textsf{index} parameter corresponds to
the current resource index and the \textsf{size} is the total number of
resources used.

@@

The server provides a number of equalizers, which automatically update
certain parameters of compounds based on runtime information. They
balance the load of 2D and DB compounds, optimally distribute render
resources for segment projections systems, adjust the resolution
to provide a constant framerate or zoom images to allow monitoring of
another view.

@@

\subsection{Name Resolution}
Pay attention that your name resolution works properly, that is, the output of
the hostname command should resolve to the same, non-local IP address on all
machines. If this is not the case, you will have to provide the public IP
address for all processes, in the following way:
\begin{description}
\item[Server] Specify the server IP as the hostname field in the connection
  description in the server section.
\item[Application] Start the application using the IP as the listening port,
  e.g., \textsf{./build/Linux/bin/eqPly -- --eq-listen 192.168.0.1}
\item[Render Clients] Specify the client IP's as the hostname field in the
  connection description of each node section.
\end{description}

@@

\begin{itemize}
\item Set the \textbf{connection} hostname and port parameters of each node in
  the configuration file.
\item Start the \textbf{render clients} using the parameters
  \textsf{--eq-client} and \textsf{--eq-listen}, e.g.,
  \textsf{./build/Linux/bin/eqPly -- --eq-client --eq-listen
  192.168.0.2:1234}. Pay attention to use the same connection parameters as in
    the configuration file.
\item Start the \textbf{application}. If the server is running on the same
  machine and user as the application, the application will connect to it
  automatically. Otherwise use the \textsf{--eq-server} parameter to specify the
  server address.
\end{itemize}

@@

{\footnotesize\begin{lstlisting}
void Channel::frameDraw( const uint32_t spin )
{
    // setup OpenGL State
    eq::Channel::frameDraw( spin );

    const float lightPos[] = { 0.0f, 0.0f, 1.0f, 0.0f };
    glLightfv( GL_LIGHT0, GL_POSITION, lightPos );

    const float lightAmbient[] = { 0.2f, 0.2f, 0.2f, 1.0f };
    glLightfv( GL_LIGHT0, GL_AMBIENT, lightAmbient );

    // rotate scene around the origin
    glRotatef( static_cast< float >( spin ) * 0.5f, 1.0f, 0.5f, 0.25f );

    // render six axis-aligned colored quads around the origin
    [...]
}
\end{lstlisting}}

@@

The API documentation on the website or in the header files provides a
comprehensive documentation on individual methods, types and other elements of
the API. Methods marked with a specific version are part of the official, public
API and have been introduced by this Equalizer version. Reasonable care is taken
to not break API compatibility or change the semantics of these methods within
future Equalizer versions of the same major revision. Any eventual changes to
the public is documented in the release notes.

@@

\begin{description}
\item[eq] The core namespace for the Equalizer client library. The
  classes and their relationship in this namespace closely model the
  configuration file format. The classes in the \textsf{eq} namespace
  are the main subject of this Programming Guide.
\item[eq::util] The eq::util namespace provides common utility classes, which
  often simplify the usage of OpenGL functions. Most classes in this namespace
  are used by the the Equalizer client library, but are usable independently
  from Equalizer for application development.
\item[eq::fabric] The eq::fabric namespace is the common data management and
  transport layer between the client library, the server and the upcoming
  administrative API. Most Equalizer client classes inherit from base classes in
  this namespace and have all their data access methods in these base
  classes. Furthermore, this namespace contains a \textsf{Serializable} class
  which implements one common use pattern of object-based data distribution
  using \textsf{eq::net::Object}.
\item[eq::net] The Equalizer network layer provides basic functionality
  for network communication, such as \textsf{Connection} and
  \textsf{ConnectionSet}, as well as higher-level functionality such as
  \textsf{Node}, \textsf{Session} and \textsf{Object}. Please refer to
  \sref{sNetwork} for an introduction into the network layer, and to
  \sref{sNetObject} for distributed objects.
\item[eq::base] The eq::base namespace provides C++ classes to abstract
  the underlying operating system and implements common helper
  functionality. One example is the \textsf{eq::base::Clock} class
  providing a high-resolution timer. Classes in this namespace are fully
  documented in the API documentation on the Equalizer website, and are
  not subject of this Programming Guide.
\end{description}

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/namespaces}
  {\caption{\label{fNamespaces}Namespace Layering for eqPly and eqServer}}
\end{wrapfigure}
Examples shipped with Equalizer are implemented in their own namespaces, e.g.,
\textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing classes from
the \textsf{eq} namespace, with the occasional usage of functionality from the
\textsf{eq::util}, \textsf{eq::fabric}, \textsf{eq::net} and \textsf{eq::base}
namespace.

The functionality of the Equalizer server is implemented in the
\textsf{eq::server} namespace. The \textsf{eq::admin} namespace enable run-time
administration of Equalizer servers. These namespace does not yet expose an API
intended to be used by developers. \fig{fClientUml} provides an overview map of
the most important classes in the Equalizer namespace, grouped by functionality.

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/clientUML}
  {\caption{\label{fClientUml}Equalizer client UML map}}
\end{figure}

@@

The distributed data classes are helper classes based on
\textsf{eq::fabric::Serializable} or its base class
\textsf{eq::net::Object}. They illustrate the typical usage of distributed
objects for static as well as dynamic, frame-specific data. Furthermore they are
used for a basic scene graph distribution of the model data.

@@

The main function starts off with parsing the command line into the
\textsf{LocalInitData} data structure. A part of it, the base class
\textsf{InitData}, will be distributed to all render client nodes. The
command line parsing is done by the \textsf{LocalInitData} class, which
is discussed in \sref{sInitData}:

{\footnotesize\begin{lstlisting}
int main( int argc, char** argv )
{
    // 1. parse arguments
    eqPly::LocalInitData initData;
    initData.parseArguments( argc, argv );
\end{lstlisting}}

The second step is to redirect all Equalizer output to a file if this
was requested through the command line. This should happen before
initializing Equalizer, so that output during initialization is
redirected and threads launched by Equalizer will use the given log
stream. Any \textsf{std::ostream} can be used as log output:

{\footnotesize\begin{lstlisting}
    // 2. Redirect log output to file, if requested
    std::ofstream* logFile = 0;
    const std::string& logFilename = initData.getLogFilename();
    if( !logFilename.empty( ))
    {
        logFile = new ofstream( logFilename.c_str( ));
        eq::base::Log::setOutput( *logFile );
    }
\end{lstlisting}}

The third step is to initialize the Equalizer library. The
initialization function of Equalizer also parses the command line, which
is used to set certain default values based on Equalizer-specific
options\footnote{Equalizer-specific options always start with -\,-eq-},
e.g., the default server address. Furthermore, a \textsf{NodeFactory} is
provided. The \textsf{EQERROR} macro, and its counterparts
\textsf{EQWARN}, \textsf{EQINFO} and \textsf{EQVERB} allow selective
debugging outputs with various logging levels:

{\footnotesize\begin{lstlisting}
    // 2. Equalizer initialization
    NodeFactory nodeFactory;
    if( !eq::init( argc, argv, &nodeFactory ))
    {
        EQERROR << "Equalizer init failed" << endl;
        return EXIT_FAILURE;
    }
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
class NodeFactory : public eq::NodeFactory
{
public:
    virtual eq::Config*  createConfig( eq::ServerPtr parent )
        { return new eqPly::Config( parent ); }
    virtual eq::Node*    createNode( eq::Config* parent )  
        { return new eqPly::Node( parent ); }
    virtual eq::Pipe*    createPipe( eq::Node* parent )
        { return new eqPly::Pipe( parent ); }
    virtual eq::Window*  createWindow( eq::Pipe* parent )
        { return new eqPly::Window( parent ); }
    virtual eq::Channel* createChannel( eq::Window* parent )
        { return new eqPly::Channel( parent ); }
    virtual eq::View* createView() { return new eqPly::View(); }
};
\end{lstlisting}}

The fourth step is to create an instance of the application and to
initialize it locally. The application is a subclass of
\textsf{eq::Client}, which in turn is an \textsf{eq::net::Node}. The
underlying Equalizer network layer, discussed in \sref{sNetwork}, is a
peer-to-peer network of \textsf{eq::net::Node}s. The client/server
concept is implement the higher-level \textsf{eq} client namespace.

The local initialization of a node creates at least one local listening
socket, which allows the \textsf{eq::Client} to communicate over the
network with other nodes, such as the server and the rendering
clients. The listening socket(s) can be configured using the
\textsf{--eq-listen} command line parameter, or by programmatically
adding connection descriptions to the client before the local
initialization.

{\footnotesize\begin{lstlisting}
    // 4. initialization of local client node
    RefPtr< eqPly::Application > client = new eqPly::Application( initData );
    if( !client->initLocal( argc, argv ))
    {
        EQERROR << "Can't init client" << endl;
        eq::exit();
        return EXIT_FAILURE;
    }
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
    // 5. run client
    const int ret = client->run();
\end{lstlisting}}

@@

{\footnotesize\begin{lstlisting}
    // 6. cleanup and exit
    client->exitLocal();
    client = 0;

    eq::exit();

    if( logFile )
        logFile->close();
    delete logFile;

    return ret;
}
\end{lstlisting}}

@@

Another configuration can be selected using the command-line parameter
\textsf{--eq-config}. This configuration may only contain an
\textsf{appNode}, i.e., no distributed configuration is possibly since
the process-local server does not create a listening socket.

{\footnotesize\begin{lstlisting}
int Application::run()
{
    // 1. connect to server
    eq::ServerPtr server = new eq::Server;
    if( !connectServer( server ))
    {
        EQERROR << "Can't open server" << endl;
        return EXIT_FAILURE;
    }
\end{lstlisting}}%>>

The second step is to ask the server for a configuration. The
\textsf{ConfigParams} are a placeholder for later Equalizer
implementations to provide additional hints and information to the
server for choosing the configuration. The configuration chosen by the
server is created locally using
\textsf{NodeFactory::createConfig}. Therefore it is of type
\textsf{eqPly::Config}, but the return value is \textsf{eq::Config},
making the cast necessary:

{\footnotesize\begin{lstlisting}
    // 2. choose config
    eq::ConfigParams configParams;
    Config* config = static_cast<Config*>(server->chooseConfig( configParams ));

    if( !config )
    {
        EQERROR << "No matching config on server" << endl;
        disconnectServer( server );
        return EXIT_FAILURE;
    }
\end{lstlisting}}%>>

Finally it is time to initialize the configuration. For statistics, the
time for this operation is measured and printed. During initialization
the server launches and connects all render client nodes, and calls the
appropriate initialization task methods, as explained in later
sections. \textsf{Config::init} returns after all nodes, pipes,
windows and channels are initialized. It returns \textsf{true} only if
all initialization task methods were successful.

@@

{\footnotesize\begin{lstlisting}
    // 3. init config
    eq::base::Clock clock;

    config->setInitData( _initData );
    if( !config->init( ))
    {
        server->releaseConfig( config );
        disconnectServer( server );
        return EXIT_FAILURE;
    }

    EQLOG( eq::LOG_STATS ) << "Config init took " << clock.getTimef() << " ms"
                           << endl;
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
    // 4. run main loop
    uint32_t maxFrames = _initData.getMaxFrames();

    clock.reset();
    while( config->isRunning( ) && maxFrames-- )
    {
        config->startFrame();
        // config->renderData(...);
        config->finishFrame();

        while( !config->needsRedraw( )) // wait for an event requiring redraw
        {
            if( hasCommands( )) // execute non-critical pending commands
            {
                processCommand();
                config->handleEvents(); // non-blocking
            }
            else  // no pending commands, block on user event
            {
                const eq::ConfigEvent* event = config->nextEvent();
                if( !config->handleEvent( event ))
                    EQVERB << "Unhandled " << event << endl;
            }
        }
        config->handleEvents(); // process all pending events
    }
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
    const uint32_t frame = config->finishAllFrames();
    const float    time  = clock.getTimef();
    EQLOG( eq::LOG_STATS ) << "Rendering took " << time << " ms (" << frame
                           << " frames @ " << ( frame / time * 1000.f)
                           << " FPS)" << endl;
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
    // 5. exit config
    clock.reset();
    config->exit();
    EQLOG( eq::LOG_STATS ) << "Exit took " << clock.getTimef() << " ms" <<endl;

    // 6. cleanup and exit
    server->releaseConfig( config );
    if( !disconnectServer( server ))
        EQERROR << "Client::disconnectServer failed" << endl;
    server = 0;
    return EXIT_SUCCESS;
}
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
bool Application::clientLoop()
{
    if( !_initData.isResident( )) // execute only one config run
        return eq::Client::clientLoop();

    // else execute client loops 'forever'
    while( true ) // TODO: implement SIGHUP handler to exit?
    {
        if( !eq::Client::clientLoop( ))
            return false;
        EQINFO << "One configuration run successfully executed" << endl;
    }
    return true;
}
\end{lstlisting}}%>>

@@

Equalizer provides distributed objects which facilitate the implementation of
data distribution in a cluster environment. Distributed objects are created by
subclassing from \textsf{eq::fabric::Serializable} or
\textsf{eq::net::Object}. The application programmer implements serialization
and deserialization of the distributed data. \sref{sNetObject2} covers
distributed objects in detail.

@@

The data streams implement efficient buffering and data transport
between nodes. They perform no type checking or transformation on the
data. It is the application's responsibility to exactly match the order
and types of variables during serialization and de-serialization.

@@

{\footnotesize\begin{lstlisting}
void InitData::getInstanceData( eq::net::DataOStream& os )
{
    os << _frameDataID << _windowSystem << _renderMode << _useGLSL << _invFaces
       << _logo;
}

void InitData::applyInstanceData( eq::net::DataIStream& is )
{
    is >> _frameDataID >> _windowSystem >> _renderMode >> _useGLSL >> _invFaces
       >> _logo;

    EQASSERT( _frameDataID != EQ_ID_INVALID );
    EQINFO << "New InitData instance" << endl;
}
\end{lstlisting}}%>>

>>>>

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications using
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.7.10 for Equalizer 1.0$\alpha$ \\[\medskipamount]
%  \today
}

@@

\lstset{language=C++}
\section*{Equalizer 1.0$\alpha$ Programming and User Guide}
\today
\subsection*{Contributors}

@@

The images on the front page show the following applications build using
Equalizer: 
RTT DeltaGen\footnote{Image copyright Realtime Technology AG, 2008} [top
right], 
\textsf{Bino}, a stereo-capable, multi-display video player\footnote{\link{http://bino.nongnu.org/}}
[middle center],
a flow visualization application for climate research\footnote{Image
  courtesy of Computer Graphics and Multimedia Systems, University of
  Siegen} [middle right],
the \textsf{eqPly} polygonal renderer in a three-sided CAVE [bottom
left],
the \textsf{eVolve} volume renderer\footnote{Data set courtesy of
  General Electric, USA} [bottom center],
and the \textsf{eqNBody} gravitational n-body simulation, based on the
N-Body CUDA example [bottom right].

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
    1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
    1.7.2   & Nov 18, 2009 & Add \sref{sSubpixel} and \sref{sFSAA} for Subpixel compounds\\
    1.7.3   & Feb 02, 2010 & Revision for Equalizer 0.9.1\\
    1.7.4   & Feb 22, 2010 & Add \sref{sSlaveCommit}\\
    1.7.5   & Jun 03, 2010 & Add \sref{sClusterSetup}\\
    1.7.6   & Aug 23, 2010 & General update for fabric changes\\
    1.7.7   & Oct 01, 2010 & SystemPipe and SystemWindow renaming\\
    1.7.8   & Oct 04, 2010 & Document GPU affinity workaround hint\\
    1.7.9   & Nov 03, 2010 & Add \sref{sErrorHandling}\\
    1.7.10  & Dec 21, 2010 & Review for Equalizer 1.0$\alpha$\\
    \hline
  \end{tabularx}}
\clearpage

@@

Equalizer is the standard middleware for the development and deployment of
parallel OpenGL applications. It enables applications to benefit from multiple
graphics cards, processors and computers to improve the rendering performance,
visual quality and display size. An Equalizer-based application runs unmodified
on any visualization system, from a simple workstation to large scale graphics
clusters, multi-GPU workstations and Virtual Reality installations.

@@

Equalizer provides the domain-specific parallel rendering know-how and abstracts
configuration, threading, synchronization, windowing and event handling. It is a
`GLUT on steroids', providing parallel and distributed execution, scalable
rendering features, an advanced network library and fully customizable event
handling.

@@

Equalizer uses CMake to generate platform-specific build files. Compiling
Equalizer is as simple as running \textsf{make} on Linux or Mac OS X, running
\textsf{make xcode} and building the generated XCode project on Mac OS X, or
creating a Equalizer Visual Studio solution using \textsf{CMake} on Windows.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/processes.pdf}
  {\caption{\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The Equalizer architecture is based on a client-server model. The client
library exposes all functionality discussed in this document to the
programmer, and provides communication between the different Equalizer
processes.

Collage, the underlying network library, uses a peer-to-peer
structure. Equalizer utilizes this to provide optimal performance scalable
rendering, e.g., by setting up direct communication between two nodes when
needed for image compositing or software swap barriers.

@@

The application can also be a rendering client, in which case it can
also contribute to the rendering. If it does not implement any render
client code, it is reduced to be the application's `master' process
without any OpenGL windows and 3D rendering.

@@

Stereo compounds, as shown in \fig{fStereoCmp}\footnote{3D model
  courtesy of Stereolithography Archive at Clemson University.}, assign
each eye pass to individual rendering units. The resulting images are
copied to the appropriate stereo buffer. This mode supports a variety of
stereo modes, including active (quad-buffered) stereo, anaglyphic stereo
and auto-stereo displays with multiple eye passes.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
  {\caption{\label{fStereoCmp}\small A Stereo Compound}}
\end{wrapfigure}
Stereo compounds are used by all applications when rendering in stereo,
and is often combined with other modes.

@@

Applications can use subpixel compounds to accelerate depth-of-field effects and
software anti-aliasing, with potentially a different number of idle or non-idle
samples per pixel.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Subpixel.pdf}
  {\caption{\label{fSubpixel}\small A Subpixel Compound}}
\end{wrapfigure}
Subpixel compounds are configured using the \textsf{subpixel} parameter, using
the values \textsf{[ index size ]}. The \textsf{index} parameter corresponds to
the current resource index and the \textsf{size} is the total number of
resources used.

@@

The server provides a number of equalizers, which automatically update
certain parameters of compounds based on runtime information. They
balance the load of 2D and DB compounds, optimally distribute render
resources for multi-display projection systems, adjust the resolution
to provide a constant framerate or zoom images to allow monitoring of
another view.

@@

\subsection{Name Resolution}
Pay attention that your name resolution works properly, that is, the output of
the hostname command should resolve to the same, non-local IP address on all
machines. If this is not the case, you will have to provide the public IP
address for all processes, in the following way:
\begin{description}
\item[Server] Specify the server IP as the hostname field in the connection
  description in the server section.
\item[Application] Specify the application IP as the hostname field in the
  connection description in the appNode section. If the application should not
  contribute to the rendering, set up an appNode without a pipe section.
\item[Render Clients] Specify the client IP's as the hostname field in the
  connection description of each node section.
\end{description}

@@

\begin{itemize}
\item Set the \textbf{connection} hostname and port parameters of each node in
  the configuration file.
\item Start the \textbf{render clients} using the parameters
  \textsf{--eq-client} and \textsf{--eq-listen}, e.g.,
  \textsf{./build/Linux/bin/eqPly --eq-client --eq-listen
  192.168.0.2:1234}. Pay attention to use the same connection parameters as in
    the configuration file.
\item Start the \textbf{application}. If the server is running on the same
  machine and user as the application, the application will connect to it
  automatically. Otherwise use the \textsf{--eq-server} parameter to specify the
  server address.
\end{itemize}

@@

{\footnotesize\lstinputlisting[firstline=107, lastline=123]
  {../../../src/examples/eqHello/hello.cpp}}

@@

The API documentation on the website or in the header files provides a
comprehensive documentation on individual methods, types and other elements of
the API. Methods marked with a specific version are part of the official, public
API and have been introduced by this Equalizer version. Reasonable care is taken
to not break API compatibility or change the semantics of these methods within
future Equalizer versions of the same major revision. Any eventual changes to
the public API is documented in the release notes and the file
\textsf{CHANGES.txt}.

@@

\begin{description}
\item[eq] The core namespace for the Equalizer client library. The
  classes and their relationship in this namespace closely model the
  configuration file format. The classes in the \textsf{eq} namespace
  are the main subject of this Programming Guide.
\item[eq::util] The eq::util namespace provides common utility classes, which
  often simplify the usage of OpenGL functions. Most classes in this namespace
  are used by the the Equalizer client library, but are usable independently
  from Equalizer for application development.
\item[eq::fabric] The eq::fabric namespace is the common data management and
  transport layer between the client library, the server and the administrative
  API. Most Equalizer client classes inherit from base classes in this namespace
  and have all their data access methods in these base classes. Furthermore,
  this namespace contains a \textsf{Serializable} class which implements one
  common use pattern of object-based data distribution using
  \textsf{co::Object}.
\item[co] Collage is the network library used by Equalizer. It provides basic
  functionality for network communication, such as \textsf{Connection} and
  \textsf{ConnectionSet}, as well as higher-level functionality such as
  \textsf{Node}, \textsf{LocalNode} and \textsf{Object}. Please refer to
  \sref{sNetwork} for an introduction into the network layer, and to
  \sref{sNetObject} for distributed objects.
\item[co::base] The co::base namespace provides C++ classes to abstract
  the underlying operating system and implements common helper
  functionality. One example is the \textsf{co::base::Clock} class
  providing a high-resolution timer. Classes in this namespace are fully
  documented in the API documentation on the Equalizer website, and are
  not subject of this Programming Guide.
\end{description}

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/namespaces}
  {\caption{\label{fNamespaces}Namespace Layering for eqPly and eqServer}}
\end{wrapfigure}
Examples shipped with Equalizer are implemented in their own namespaces, e.g.,
\textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing classes from
the \textsf{eq} namespace, with the occasional usage of functionality from the
\textsf{eq::util}, \textsf{eq::fabric}, \textsf{co} and \textsf{co::base}
namespace.

The functionality of the Equalizer server is implemented in the
\textsf{eq::server} namespace. The \textsf{eq::admin} namespace enables run-time
administration of Equalizer servers. These namespaces do not yet expose a stable
API. \fig{fClientUml} provides an overview map of the most important classes in
the Equalizer namespace, grouped by functionality.

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/clientUML}
  {\caption{\label{fClientUml}Equalizer client UML map}}
\end{figure}

@@

The distributed data classes are helper classes based on
\textsf{eq::fabric::Serializable} or its base class \textsf{co::Object}. They
illustrate the typical usage of distributed objects for static as well as
dynamic, frame-specific data. Furthermore they are used for a basic scene graph
distribution of the model data.

@@

The main function starts off with initializing the Equalizer library. The
command line arguments are passed on to Equalizer. The are used to set certain
default values based on Equalizer-specific options\footnote{Equalizer-specific
  options always start with -\,-eq-}, e.g., the default server
address. Furthermore, a \textsf{NodeFactory} is provided. The \textsf{EQERROR}
macro, and its counterparts \textsf{EQWARN}, \textsf{EQINFO} and \textsf{EQVERB}
allow selective debugging outputs with various logging levels:

{\footnotesize\lstinputlisting[firstline=57, lastline=66]
  {../../../src/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=40, lastline=55]
  {../../../src/examples/eqPly/main.cpp}}

The second step is to parse the command line into the
\textsf{LocalInitData} data structure. A part of it, the base class
\textsf{InitData}, will be distributed to all render client nodes. The
command line parsing is done by the \textsf{LocalInitData} class, which
is discussed in \sref{sInitData}:

{\footnotesize\lstinputlisting[firstline=67, lastline=70]
  {../../../src/examples/eqPly/main.cpp}}

The third step is to create an instance of the application and to
initialize it locally. The application is a subclass of
\textsf{eq::Client}, which in turn is an \textsf{co::LocalNode}. The
underlying Collage network library, discussed in \sref{sNetwork}, is a
peer-to-peer network of \textsf{co::LocalNode}s. The client/server
concept is implement the higher-level \textsf{eq} client namespace.

The local initialization of a node creates at least one local listening socket,
which allows the \textsf{eq::Client} to communicate over the network with other
nodes, such as the server and the rendering clients. The listening socket(s) can
be configured using the \textsf{--eq-listen} command line parameter, by adding
connections to the \textsf{appNode} in the configuration file, or by
programmatically adding connection descriptions to the client before the local
initialization:

{\footnotesize\lstinputlisting[firstline=71, lastline=78]
  {../../../src/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=80, lastline=81]
  {../../../src/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=83]
  {../../../src/examples/eqPly/main.cpp}}

@@

Another configuration can be selected using the command-line parameter
\textsf{--eq-config}. If this configuration contains additional nodes, the
\textsf{server} and \textsf{appNode} sections need a globally reachable
\textsf{connection} configured for the render clients to connect to.

{\footnotesize\lstinputlisting[firstline=80, lastline=88]
  {../../../src/examples/eqPly/eqPly.cpp}}

The second step is to ask the server for a configuration. The
\textsf{ConfigParams} are a placeholder for later Equalizer
implementations to provide additional hints and information to the
server for choosing the configuration. The configuration chosen by the
server is created locally using
\textsf{NodeFactory::createConfig}. Therefore it is of type
\textsf{eqPly::Config}, but the return value is \textsf{eq::Config},
making the static\_cast necessary:

{\footnotesize\lstinputlisting[firstline=90, lastline=101]
  {../../../src/examples/eqPly/eqPly.cpp}}

Finally it is time to initialize the configuration. For statistics, the
time for this operation is measured and printed. During initialization
the server launches and connects all render client nodes, and calls the
appropriate initialization task methods, as explained in later
sections. \textsf{Config::init} returns after all nodes, pipes,
windows and channels are initialized.

The return value of \textsf{Config::init} depends on the configuration
robustness attribute. This attribute is by default activated, allowing
configurations to launch even when some entities failed to initialize. If
activated, \textsf{Config::init} always returns true. If deactivated, it returns
\textsf{true} only if all initialization task methods were successful. In any
case, \textsf{Config::getError} only returns \textsf{ERROR\_NONE} if all
entities have initialized successfully.

@@

{\footnotesize\lstinputlisting[firstline=103, lastline=118]
  {../../../src/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=120, lastline=147]
  {../../../src/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=148, lastline=152]
  {../../../src/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=154, lastline=169]
  {../../../src/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=171, lastline=179]
  {../../../src/examples/eqPly/eqPly.cpp}}

@@

Equalizer provides distributed objects which facilitate the implementation of
data distribution in a cluster environment. Distributed objects are created by
subclassing from \textsf{eq::fabric::Serializable} or \textsf{co::Object}. The
application programmer implements serialization and deserialization of the
distributed data. \sref{sNetObject2} covers distributed objects in detail.

@@

The data streams implement efficient buffering and compression, and
automatically select the best connection, i.e., multicast where available, for
data transport. They perform no type checking or transformation on the data. It
is the application's responsibility to exactly match the order and types of
variables during serialization and de-serialization.

@@

{\footnotesize\lstinputlisting[firstline=61, lastline=74]
  {../../../src/examples/eqPly/initData.cpp}}
